Compare NE-combined vs. NE-separated reference

### 2025-07-09 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/separate-tumor.Rmd
UMAP and update to expression plot

We'll begin with a heatmap directly comparing annotations between the two SingleR models:

@@

### Cells labeled as Neuroendocrine or Neuroendocrine-tumor

We'll first directly compare cells that were labeled as a `Neuroendocrine` cell type with either model.
Each row shows counts of cells that had given pairs corresponding labels from inference with either model.

```{r}
singler_df |>
  dplyr::count(singler_label_combined, singler_label_separated) |>
  dplyr::arrange(desc(n))
```

### Unlabeled cells

@@

Next, we'll pull out the _distinct_ marker genes associated with `NE` and `NE-tumor` groups that are also not represented in the combined model's `NE` marker genes.

@@

```{r}
ne_marker_genes <- model_separated$markers$full$Neuroendocrine |>
  purrr::reduce(c) |>
  unique()

@@

ne_only <- setdiff(ne_marker_genes, ne.tumor_marker_genes)
ne.tumor_only <- setdiff(ne.tumor_marker_genes, ne_marker_genes)
cell_types <- names(ne_markers)
non_neuro_cell_types <- cell_types[!(stringr::str_detect(cell_types, "Neuro"))]

different_marker_genes <- non_neuro_cell_types |>
  purrr::set_names() |>
  purrr::map(
    \(cell_type) {
      # In NE but not in NE-tumor
      diff_genes_ne <- setdiff(
        ne_markers[[cell_type]],
        ne.tumor_markers[[cell_type]]
      )

      # In NE-tumor but not in NE
      diff_genes_ne.tumor <- setdiff(
        ne.tumor_markers[[cell_type]],
        ne_markers[[cell_type]]
      )

      return(
        list(
          "ne_only" = diff_genes_ne,
          "ne.tumor_only" = diff_genes_ne.tumor
        )
      )
    }
  ) |>
  # transpose so it is a list of two, "ne_only" and "ne.tumor_only",
  # each of which is a named list of marker genes for each cell type
  purrr::transpose() |>
  # collapse each nested list to remove cell types
  # keep only unique genes
  # remove any genes that are also in combined_marker_genes
  purrr::map(
    \(gene_list) {
      gene_vector <- gene_list |>
        purrr::reduce(c) |>
        unique()
      gene_vector[!(gene_vector %in% combined_marker_genes)]
    }
  )

# This final list are the genes uniquely upregulated in the NE and
# NE-tumor categories in the separated model
different_marker_genes
```

```{r, message=F}
plot_barcodes <- singler_df |>
  dplyr::filter(stringr::str_detect(singler_label_separated, "Neuroendocrine")) |>
  dplyr::pull(barcodes)

@@

```{r, fig.width = 12, message=F, warning=F}
# plot the NE-only genes
plot1 <- scater::plotExpression(
  sce[different_marker_genes$ne_only, plot_barcodes],
  features = different_marker_genes$ne_only,
  x = "singler_cell_type",
  colour_by = "singler_cell_type",
  point_size = 0.5,
) +
  stat_summary(shape = 18) +
  theme(legend.position = "none") +
  ggtitle("Neuroendocrine marker genes")

# plot the NE-tumor--only genes
plot2 <- scater::plotExpression(
  sce[different_marker_genes$ne.tumor_only, plot_barcodes],
  features = different_marker_genes$ne.tumor_only,
  x = "singler_cell_type",
  colour_by = "singler_cell_type",
  point_size = 0.5
) +
  stat_summary(shape = 18) +
  theme(legend.position = "none") +
  ggtitle("Neuroendocrine-tumor marker genes")

plot1 + plot2
```

>>>>

### UMAP

```{r}
# colors adapted from:
# https://github.com/AlexsLemonade/scpca-paper-figures/blob/a457408323495b6ecee0c63ccdd74b41e88ab703/palettes/validation-group-palette.tsv
nbatlas_colors <- c(
  "B cell" = "#AA0DFE",
  "Endothelial" = "#FC1CBF",
  "Fibroblast" = "#B10DA1",
  "Myeloid" = "#90AD1C",
  "Neuroendocrine" = "#FEAF16",
  "Neuroendocrine-tumor" = "#B17A0F",
  "NK cell" = "#FBE426",
  "pDC" = "#16FF32",
  "Plasma" = "#782AB6",
  "RBCs" = "#F6222E",
  "Schwann" = "#1CFFCE",
  "Stromal other" = "#C4451C",
  "T cell" = "#3283FE"
)
singler_long_df <- singler_df |>
  tidyr::pivot_longer(
    contains("singler_label"),
    names_to = "model",
    values_to = "singler_label"
  ) |>
  dplyr::mutate(model = stringr::str_remove(model, "singler_label_"))
```

First, we'll show all cell types:
```{r}
# remove NAs before plotting
ggplot(tidyr::drop_na(singler_long_df)) +
  aes(x = UMAP1, y = UMAP2, color = singler_label) +
  geom_point(size = 0.5, alpha = 0.5) +
  scale_color_manual(values = nbatlas_colors) +
  facet_wrap(vars(model)) +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(override.aes = list(size = 1, alpha = 1)))
```

Next, we'll show this faceted only for the `NE` cell types
```{r}
singler_ne_df <- singler_long_df |>
  dplyr::filter(stringr::str_detect(singler_label, "Neuroendocrine"))

# remove NAs before plotting
ggplot(singler_ne_df) +
  aes(x = UMAP1, y = UMAP2, color = singler_label) +
  geom_point(size = 0.75, alpha = 0.5) +
  scale_color_manual(values = nbatlas_colors) +
  facet_grid(
    rows = vars(model),
    cols = vars(singler_label)
  ) +
  theme(legend.position = "none")
```

### Jaccard similarity heatmap

Next we'll show a heatmap directly comparing annotations between the two SingleR models colored by Jaccard similarity.

@@

#### All labels

The follow table shows relationship between model annotations:

```{r}
table(
  singler_df$singler_label_combined,
  singler_df$singler_label_separated
)
```

#### Unlabeled cells

@@

Next, we'll pull out the _distinct_ marker genes associated with `NE` and `NE-tumor` groups.

@@

```{r}
# Define the marker genes
ne_marker_genes <- model_separated$markers$full$Neuroendocrine |>
  purrr::reduce(c) |>
  unique()

@@

ne_only <- setdiff(ne_marker_genes, ne.tumor_marker_genes)
ne.tumor_only <- setdiff(ne.tumor_marker_genes, ne_marker_genes)
```

```{r, message=F}
neuro_barcodes <- singler_df |>
  dplyr::filter(stringr::str_detect(singler_label_separated, "Neuroendocrine")) |>
  dplyr::pull(barcodes)

@@

```{r, fig.width = 14, message=F, warning=F}
# plot the NE-only genes
plot1 <- scater::plotExpression(
  sce[ne_only, neuro_barcodes],
  features = ne_only,
  x = "singler_cell_type",
  colour_by = "singler_cell_type",
  point_size = 0.5,
  ncol = 6 # facet cols
) +
  stat_summary(shape = 18) +
  theme(legend.position = "none") +
  ggtitle("Neuroendocrine marker genes")

# plot the NE-tumor--only genes
plot2 <- scater::plotExpression(
  sce[ne.tumor_only, neuro_barcodes],
  features = ne.tumor_only,
  x = "singler_cell_type",
  colour_by = "singler_cell_type",
  point_size = 0.5,
  ncol = 6 # facet cols
) +
  stat_summary(shape = 18) +
  theme(legend.position = "none") +
  ggtitle("Neuroendocrine-tumor marker genes")

plot1 / plot2
```
